Collaborative Multioutput Gaussian Processes

### 2014-03-16 paper/inference.tex
latest experiment results and paper style
model description and inference

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\

\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
\begin{align}
\nonumber
p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
\nonumber
&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
\label{eq:sorprediction}
&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
\end{align}
where,
\begin{align}
\nonumber
\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
\nonumber
\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
\end{align}
 Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.

\subsection{Variational Inference}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our goal of inference is to find the posterior $p(\g, \h, \u, \v | \y)$. 
Following the previous discussion, we assume a variational distribution which factorizes as:
\begin{align}
\nonumber
q(\g, \h, \u, \v | \y)
\nonumber
 &= p(\g|\u) p(\h|\v) q(\u,\v)  \\
 &= p(\g|\u) p(\h|\v) \prod_{j=1}^Q q(\u_j) \prod_{i=1}^P  q(\v_i)
\end{align}
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only  find the optimum $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = q(\v_i) = \Normal(\v_i; \mhi, \Shi)$.

@@

\subsubsection{Variational Parameters Derivatives}
\newcommand{\oi}{\vec{o}_i}
% some notation
Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
Readers familiar with this operator will recognize that this is the MATLAB indexing operator.

@@

%\subsubsection{Update Equations}
% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
\subsection{Prediction}
The predictive distribution of the $i$-th output for a test input $\x_*$ is 
\begin{align}
p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
\end{align}
where $p(\g_* | \y, \x_*) = \prod_{j=1}^Q \Normal(g_{j*}; \mu_{j*}, s_{j*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu^h_{i*}, s^h_{i*})$ are the predictive distributions of the sparse GPs as given in eq. \ref{eq:sorprediction}.
Therefore we have:
\begin{align}
p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu_{i*}, w_{ij}^2 s_{j*} + s_{i*}). 
\end{align}

>>>>

%\subsection{Variational Inference in Sparse GPs Revisited}
%
%In this section we review inference in sparse GPs (as presented in Titsias).
%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
%
%\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
%\begin{align}
%\nonumber
%p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
%\nonumber
%&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
%&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
%\label{eq:sorprediction}
%&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
%\end{align}
%where,
%\begin{align}
%\nonumber
%\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
%\nonumber
%\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
%\end{align}
% Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.

Inference in this model is done with variational inference.
In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the novel property of factorizing over the data points.
The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates the use of a distribution where the true posteriors over $\g$ and $\h$ are replaced by their conditional distributions given the inducing variables:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used in \citep{titsiasvariational,hensmangaussian} to derive variational inference for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (Equation \ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors, which becomes small when the sufficiency condition holds with greater certainty.

@@

\subsection{STOCHASTIC VARIATIONAL INFERENCE}
\newcommand{\oi}{\vec{o}_i}
% some notation

@@

\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
%\subsubsection{Update Equations}
% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
\subsection{PREDICTION}
The predictive distribution of the $i$-th output for a test input $\x_*$ is 
\begin{align}
p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
\end{align}
where $p(\g_* | \y, \x_*) = \prod_{j=1}^Q \Normal(g_{j*}; \mu_{j*}, s_{j*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu^h_{i*}, s^h_{i*})$ are the predictive distributions of the sparse GPs as given in eq. \ref{eq:sorprediction}.
Therefore we have:
\begin{align}
p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu_{i*}, w_{ij}^2 s_{j*} + s_{i*}). 
\end{align}

### 2014-03-16 paper/model.tex
model description and inference

We describe a multiple-output regression model.
Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
For ease of exposition, we assume that there are no missing values in any output dimension.
The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
Our model is also underpinned by this composition approach.

\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
Each (latent) output is a weighted combination of these basic processes, i.e.
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
\end{align}
In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
\end{align}
We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.

\noindent 
\textbf{Augmented sparse GPs}
Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs. The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.

%However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.

\subsection{Prior and Likelihood}
In this section we specify the prior and likelihood of this model. 
\newcommand{\Zj}{\Z_j}
\newcommand{\Zhi}{\Z^h_i}
Let the set of inducing inputs for $g_j(\x)$ and $h_i(\x)$ be $\Z_j$ and $\Z^h_i$, and the corresponding inducing points $\u_j$ and $\v_i$, respectively.
We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.

\noindent
\textbf{Prior}
The prior of the augmented model is given by:
\begin{align}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
\end{align}
where $\BigMu_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}\u$ and $\tilde{\K}_j = k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X)$ and $\BigMu^h_i, \tilde{\K}^h_i$ are similarly defined.
To avoid notational clutter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
\end{align}

>>>>

Before diving into details of the specification, we discuss the philosophy and motivation behind our collaborative multioutput model.
To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
This is achieved in the model by allowing the outputs to share multiple sets of inducing variables.
For exposition, consider a set of inducing points for a moment.
Because they serve as the sufficient statistics of the individual outputs, a pattern common to the outputs can be encapsulated by these points when learning is done jointly. 
Different patterns in the data can be captured by having  multiple set of inducing variables.
It should be evident that these variables play a double pivotal role in the model: they represent the sufficient statistics that is crucial for sparse approximation and they collaborate sharing of information across the outputs.

\newcommand{\Zj}{\Z_j}
\newcommand{\Zhi}{\Z^h_i}
Consider multioutput regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$. 
Assume $Q$ latent functions with independent Gaussian process priors, $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$, $j= 1 \hdots Q$.
We introduce the sets of \emph{inducing variables} $\u_j$, each of which comes from $g_j(\x)$, i.e., $\u_j$ contains the function values at the inducing inputs $\Z_j$, which live in the same space as $\X$.
These processes and inducing sets are shared by all of the outputs.
To accommodate greater flexibility, we further allow each output to have its own independent process $h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$ whose the sets of inducing inputs and variables are $\Z^h_i$ and $\v_i$ respectively, where $i = 1 \hdots P$.
% information is transfered via the inducing variables
We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.
For convenience, we shall assume the same number of inducing points, $M$, for all of the processes, however this is not imposed in practice.

Following the definition of the GPs and the independence of the processes, the \emph{prior} of the multioutput model can be written as:
\begin{align}
\label{eq:gu}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
\label{eq:hv}
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)), \\
 \BigMu_j &= k(\X,\Zj)k(\Zj,\Zj)^{-1}\u_j \\
\BigMu^h_i &= k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}\v_i \\
\tilde{\K}_j &= k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X) \\
\tilde{\K}^h_i &= k^h_i(\X,\X) - k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}k(\Zhi,\X).
\end{align}
In the above equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.

The \emph{likelihood} with standard iid Gaussian likelihood is given by:
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
\end{align}
which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05,seeger2005semiparametric}.
However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing outputs.
Furthermore, as we shall see in the next section, explicit representation of these variables is instrumental to scalable inference of this model.

%\textbf{Augmented sparse GPs}
%Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
%This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs. The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
%Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.
